---
title: "purrr"
author: "Fonti Kar"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introducing purrr and data

```{r}
# install.packages("purrr")
library(purrr)

# or 

library(tidyverse)
library(janitor)
library(palmerpenguins)
```
Okay, we've just gone through some function writing tips and tricks and had a play on how to get them working. 

Now, I want to introduce you all to the R package purrr. This package will allow you to use your functions and scale up dramatically. 

The main functions from purrr you will likely use is `map_` and its variants `map2`, `pmap` or `walk_` and its variants. 

`map` loves split-apply-combine problems. It is written for exactly this purpose.

Let me demonstrate first. Look inside the `data/penguin_distributions/` folder. There you can see that I have multiple `.csv` files for different species of penguins. 

## Map and its structure

I need to read these `.csv` into R and join them together into one big dataframe. 

Here I'm getting a list of file paths for the files in `data/penguin_distributions`. Using `pattern=".csv"`, the function will only display files that match the pattern. `full.names` will display the full relative file path to each file (rather than the file name)

the `map()` function expects a few things for you. Look up the help file and have a see for yourself.

`map()` needs: 
- the data, a list/vector to iterate over
- a function to apply to each element of your list/vector. 

Below, I'm applying `read_csv()` to each file path name. Notice I didn't have to supply the `()` after `read_csv.` Arguments for `read_csv` are adjusted as inputs in `map` like setting FALSE for `show_col_types`

I suggest breaking up the pile and running below step to under what is happen at each line. 

The final function `list_rbind()`, glues all the different data frames together into one big long dataframe! Yay! 

### Read in lots of data

```{r}
list.files("data/penguin_distributions", pattern=".csv", full.names = TRUE)

list.files("data/penguin_distributions", pattern=".csv", full.names = TRUE) |> 
  map(read_csv, show_col_types = FALSE) 

list.files("data/penguin_distributions", pattern=".csv", full.names = TRUE) |> 
  map(read_csv, show_col_types = FALSE) |> 
  list_rbind() # Put it all together

# Finally assigning it an object  
penguins_combined <- list.files("data/penguin_distributions", pattern=".csv", full.names = TRUE) |> 
  map(read_csv, show_col_types = FALSE) |> 
  list_rbind() |> 
  clean_names()
```
### Exercise: nrow for island

- Can you get the output of `nrow` for each `island` in the palmerpenguins::penguins dataset using `map`. (Hint: You will need `split()`). 

Once you got that going, try replacing `map` with `map_dbl` - what do you notice?

```{r, learners-purrr-e1}

```

## Anonymous functions

`map()` unlocks your power to scale up tasks in R! Let's try another example. 

Here I am splitting the data by species inside R. This means that I will be operating on the species level. 

I am going to generate a plot for each species. A histogram of `individualCount`. 

Notice how ggplot code is more complicated than the `read_csv()` above. ggplot code is split across multiple lines right? This is a good time to use a anonymous function. This is like writing a function on the fly. 

Anonymous functions are often used for quick one-liners

```{r}
penguins_by_species_list <- split(penguins_combined, ~species)

penguins_by_species_list  |> 
  map(~ ggplot(.x) + geom_histogram(aes(x = individual_count)))

penguins_by_species_list  |> 
  map(~ summary(.x))
```

### Exercise: Glimpse by year

- Can you split the `penguins_combined` data by year and get the output for `glimpse()` for each year?

```{r, learners-purrr-e2}

```

## purrr + your customised functions

Here we are fitting a linear model for every species with a function we have written ourselves. 

Tip: Test out your function first before putting it through `map()`

```{r}
# Filter data to only that with count data
data_counts <- penguins_combined |> filter(individual_count > 0)

# Write a function to fit a model
fit_model <- function(data) {
  model <- lm(log10(individual_count) ~ decimal_latitude, data = data)
  return(model)
}

# Get a unique species
pen_species <- unique(data_counts$species)

# Try it with one species
data_one_spp <- filter(data_counts, species==pen_species[1])
fit_one_spp <- fit_model(data_one_spp)

# Map our fit model function
data_counts |> 
  split(~species) |>
  map(fit_model)
```

### Exercise: Get R^2 

Try extract the R squared from the models we just fitted. You may need the`summary()` and `str()`. Alternatively, you can used `broom::glance()`

```{r, learners-purrr-e3}

```

## Map over 2 inputs

So far we have only see purrr used with one input that we are iterating over, island, species, year. What if we have 2 inputs?? Meet `map2_` a family of functions that accepts 2 iterating inputs. 

Let's say I want to create a new variable which is a combination of `island` and `species` from the palmerspenguins dataset

```{r}
penguins |> 
  mutate(penguin_island = map2_chr(species, island, ~ paste(.x, .y, sep = "_"))) |> 
  relocate(penguin_island, .after = island) # Move new variable in a nice places
```

### Exercise: Ratio between bill depth and bill length

- Try create a variable for `mass_adjusted_bill_length` by taking the ratio between `bill_length` and `body_mass`

```{r}

```

## Walking instead of saving many things

`map` functions has a cousin called `walk` that essentially does the same thing...but it returns the output **invisibly**, its typically used in conjunction with functions where the output is not relevant

Here I want to save the output of the penguins data by year as .csv files

```{r}
# Split data by year
penguins_list_by_year <- 
  penguins_combined |> 
  split(~year)

# Get the names of year
names(penguins_list_by_year) <- sort(unique(penguins_combined$year))

# Mapping
purrr::map2(penguins_list_by_year, names(penguins_list_by_year), ~ write_csv(.x, file = paste0("output/penguins_", .y, ".csv")))

# Walking2
purrr::walk2(penguins_list_by_year, names(penguins_list_by_year), ~ write_csv(.x, file = paste0("output/penguins_", .y, ".csv")))

# Alternative with iwalking 
purrr::iwalk(penguins_list_by_year, ~ write_csv(.x, file = paste0("output/penguins_", .y, ".csv")))
```

### Exercise: 

- What is different between the `map2` and `walk2` call? 
- What is different in the syntax of the `walk2` and `iwalk` call? Does the help file give you a hint?
- Using palmer penguins data, write a function that will create scatterplots between bill_length and bill depth. Include a title for each species name and then save the produced plot `ggsave()`, `walk2` or `iwalk` this over each species

```{r}

```

# Nesting and list columns

Now I'd like to introduce to a new sort of data wrangling practice and structure. We have worked with lists right? Lists are nested structures that allows you to store a lot different kinds of data. 

```{r}
example_list <- list(A = c(1:5), 
                     B = c(letters[1:5]), 
                     C = tibble(x = 6:10, 
                                y = 8:12))
```

You can access elements of a list with `$` or `[[]]``

```{r}
example_list$C

#Alternatively, you can also use `purrr::pluck()`

example_list |> pluck("C")
```

Now that we got the basics under our belt, let's add a bit more complexity to this. I want to introduce a structure called a **list column**. 

We are going to create a list column, where the columnis going hold a list in each cell using `tidyr::nest()`. 

Here, I am nesting or "tucking in" the data for each `species` of penguins from the `palmerpenguins` dataset coordinates of each species into their own tibble ad storing it into a column called `data`. Check it out: 

```{r}
nested_penguins <- pÃenguins |>
  group_by(species) |>
  nest()

nested_penguins
```

This is essentially splitting the data frame by species which we've done a few times already, but isn't it much neater?! We can see that there are 3 species in total and how many observations we have for each at a glance. 

```{r}
nested_penguins
```

Let's have a peek at some data for Adelies with `pluck` 

```{r}
nested_penguins |> 
  pluck("data", 1)
```

The benefit of using list columns is to that it works _SO_ well with `purrr` functions, especially map. They go hand in hand, especially if you want to do several operations at the same "grouping variable" that you have nested at. 

Here I am computing summary statistics

```{r}
# Calculate summary statistics for each species
summary_stats <- nested_penguins |>
  mutate(summary = map(data, ~ summarise(.x, 
                                         mean_bill_length = mean(bill_length_mm, na.rm = TRUE),
                                         mean_bill_depth = mean(bill_depth_mm, na.rm = TRUE),
                                         mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
                                         mean_body_mass = mean(body_mass_g, na.rm = TRUE))))

summary_stats

# View the summary statistics
summary_stats |>
  select(species, summary) |>
  unnest(summary)
```

### Exercise: Model fit and predict

Nest the palmerpenguins data by species and fit a linear model predicting body mass from bill length for each group. 

Bonus: Once you have done can you use the `{broom}` package to `augment` some model predictions for body mass to our data frame 

```{r, learners-purrr-e4}

  
```

